Discrete Quantum Theories and Computing

### 2018-12-17 dissertation.tex
dissertation.lyx "Chapter 1 Introduction" Updating...

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter how, classical or quantum, to attack these two tasks, they
require two different types of quantities. On one hand, the question
and answer of RSA are both discrete integers. On the other hand, even
if the quantities describing molecules are considered as continuous,
we want to argue that it is beyond our ability to decide these quantities,
including the coordinate of the wave functions, are ultimately discrete
or continuous. To better understand this issue without struggling
with their meaning in the quantum theory, it is instructive to review
how classical quantities are characterized and computed.

@@

In contrast, the non-apparently discrete quantities confused many
cultures since very beginning. For example, the ancient Chinese philosopher
Chuang-tzu (庄子) didn't believe these quantities have a minimum unit,
and expressed ``If a one-foot-long stick is cut into halves every
day, the cutting will never come to an end''\footnote{``一尺之棰，日取其半，萬世不竭。'' in Chinese \cite{Zhuangzi1999,Liu2018}.}.
As another example, Greek mathematicians in the school of Pythagoras
believed every mathematical model must be ultimately discrete, and
they in legend murdered Hippasus because he proved there is no minimum
unit between a side of a square and its diagonal. Comparing the idea
that the length of a square's diagonal is not a physical quantity
or there is no square physically, working with a mathematical model
without minimum unit seems more appealing. Since the mathematical
model is continuous, their corresponding physical quantities are naturally
assumed to be continuous. Later, these continuous quantities become
the basic component of analog computers. After logarithm was discovered,
a slide rule, also known as a slipstick, is used to compute multiplication,
division, and more complex operations as a mechanical analog computer~\cite{wiki:SlideRule}.
Charge amplifiers are used to build electrical analog computers and
compute the integration in calculus \cite{wiki:Integrator,wiki:ChargeAmplifier}.

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, the physical quantities discussed in
previous paragraph may not simply represented by real numbers. Chuang-tzu's
stick are made by molecules and cannot be cut in halves endlessly.
The input charge of an integrator must actually be an integer multiple
of the elementary charge. A prefect square cannot be draw physically.
Even we don't draw it, but merely asking whether it exists in the
space is still beyond our ability to answer because whether the space
itself is ultimately discrete or continuous is still an unsolved question.
Even if it exists, it cannot be distinguished by human from an almost
prefect square differing only by an arbitrarily small extent. Similar
situation applying to the length of a slide rule, it is beyond our
ability to decide whether its length is ultimately discrete or continuous.
Even if it is continuous, identifying which real number represents
its length requires infinite precision, but the precision in reality
is limited generally three or four significant figures~\cite{wiki:AnalogComputer}.
As the results, even if analog computers really store real numbers,
they cannot be read, write, and use to its exact value to branch the
computation. The opposite of the last condition is sometimes assumed
by real computational models. For example, Blum, Cucker, Shub, and
Smale modeled their BCSS machine to allow the operations of deciding
whether a number is greater or equal to zero over real numbers and
whether a number is exactly zero over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Therefore, the computational power of analog computers is far weaker
than their theoretical prediction.

\begin{sidewaystable}
\caption{\label{tab:Comparison}Comparison between the classical quantities
and their mathematical models. \todo{Add the quantum case, and center
entries vertically.}}

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{YcYY}
\toprule 
\addlinespace
Physical quantities & Discrete & Discrete with extreme small unit & Continuous?\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & 
Not reliably, but could be represented by a faithful discrete model when the minimum unit could be manipulated with a better technology. & 
Not reliably, because precision can never be high enough. \tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and similar
situations could potentially be applied to quantum mechanics and computation.
The physical quantity manipulated by the quantum circuit model, the
most widely used model of quantum computing, is the probability amplitudes
of states. \todo{Introduction Cleaned up until here. }Because we
are agnostic about whether probability amplitude is quantized or can
really be valued in the continuum of complex numbers, we tried to
develop two different types of models: the quantum theories and computing
over finite fields are discrete models, and the quantum interval-valued
probability measures incorporate the idea of measurement precision
into the quantum theory. To distinguish the modified quantum theories
from the usual ones, the usual ones will be called the ``conventional
quantum theory'' where necessary and it will be reviewed in Chapter~\ref{chap:Conventional-Quantum-Theory}~\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}.

@@

\begin{tabularx}{\linewidth}{YYYYYYY}
\toprule 
\addlinespace
 & Conventional & Modal & Discrete (I) & Discrete (II) & QPMFF & QIVPM\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 & Chapter~\ref{chap:Conventional-Quantum-Theory} & Sec.~\ref{modalquantum} & Sec.~\ref{discretequantumtheoryI} to \ref{sec:Geometry-of-Entangled-States} & Sec.~\ref{sec:Discrete-Quantum-Theory-(II)} & Sec.~\ref{sec:Toward-IVPM} & Chapter~\ref{chap:QIVPM}\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
States space & $\mathbb{C}^{D}$ & $\Fq^{D}$ & $\Fpp^{D}$ & Local region in $\Fpp^{D}$ & $\Fpp^{D}$ & $\mathbb{C}^{D}$\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Likelihood of events is predicted by & Real-valued probability & Possible or impossible & Possible or impossible & Cardinal probability & Real-valued probability, but no sensible Born rule & Interval-valued probability\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Expectation value & Defined & Undefined & Undefined & Undefined &  & Defined\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 &  & Sec.~\ref{modalquantumcomputing} & Sec.~\ref{discretequantumcomputingI} & Sec.~\ref{discretequantumcomputingII} &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Hadamard transformation & Included & May not be included & Included & Included &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Solve $\usat$ efficiently & Unlikely & Yes & Partially & Unlikely &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Grover search algorithm & Yes &  &  & Yes &  & \tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabularx}
\end{sidewaystable}
\todo{Beginning of Merge Left}Since conventional probability amplitude
is in the field of complex numbers, and the field structure is essential
to compute the probability from some probability amplitudes, we in
Chapter~\ref{chap:DQT=000026DQC} tried to build discrete quantum
theories and computing by considering them over finite fields.

>>>>

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter how to, classical or quantum, attack these two tasks, they
require two different types of quantities. On one hand, the question
and answer of RSA are both discrete integers. On the other hand, even
if the quantities describing molecules, like their positions, are
considered as continuous, we want to argue that it is beyond our ability
to decide whether these quantities are ultimately discrete or continuous.
To better understand this issue without struggling with the quantum
theory, it is instructive to review how classical quantities are characterized
and computed.

@@

In contrast, the non-apparently discrete quantities confused many
cultures since very beginning. For example, the ancient Chinese philosopher
Chuang-tzu (庄子) didn't believe these quantities have a minimum unit,
and expressed ``If a one-foot-long stick is cut into halves every
day, the cutting will never come to an end''\footnote{``一尺之棰，日取其半，萬世不竭。'' in Chinese \cite{Zhuangzi1999,Liu2018}.}.
As another example, Greek mathematicians in the school of Pythagoras
believed every mathematical model must be ultimately discrete, and
they in legend murdered Hippasus because he proved there is no minimum
unit between a side of a square and its diagonal. Comparing the idea
that the length of a square's diagonal is not a physical quantity
or there is no square physically, working with a mathematical model
without minimum unit seems more appealing. Since the mathematical
models are continuous, their corresponding physical quantities are
naturally assumed to be continuous. Later, these continuous quantities
become the basic component of analog computers. After logarithm was
discovered, a slide rule, also known as a slipstick, is used to compute
multiplication, division, and more complex operations as a mechanical
analog computer~\cite{wiki:SlideRule}. Charge amplifiers are used
to build electrical analog computers and compute the integration in
calculus \cite{wiki:Integrator,wiki:ChargeAmplifier}.

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, the physical quantities discussed in
the previous paragraph should not simply represented by real numbers.
Chuang-tzu's stick are made by molecules and cannot be cut in halves
endlessly. The input charge of an integrator must actually be an integer
multiple of the elementary charge. A prefect square cannot be draw
physically. Even we don't draw it, but merely asking whether it exists
in the space is still beyond our ability to answer because whether
the space itself is ultimately discrete or continuous is still an
unsolved question. Even if a prefect square exists physically, we
cannot decide whether it is really a prefect square or an almost prefect
one differing only by an arbitrarily small extent. Similar situation
applying to the length of a slide rule, it is beyond our ability to
decide whether its length is ultimately discrete or continuous. Even
if its length is continuous, identifying which real number represents
its length requires infinite precision, but the precision in reality
is limited generally three or four significant figures~\cite{wiki:AnalogComputer}.
As the results, even if analog computers really store real numbers,
they cannot be read, written, and used to branch the computation based
on their exact values. The opposite of the last statement is sometimes
assumed by real computational models. For example, Blum, Cucker, Shub,
and Smale modeled their BCSS machine to allow the operations of deciding
whether a number is greater or equal to zero over real numbers and
whether a number is exactly zero over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Therefore, the computational power of analog computers is far weaker
than the prediction based on this kind of theoretical models.

\begin{sidewaystable}
\caption{\label{tab:Comparison}Comparison between the classical quantities
and their mathematical models with their inspired quantum models.
\todo{Center entries vertically.}}

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{ccYY}
\toprule 
\addlinespace
Physical quantities & Discrete & Discrete with extreme small unit & Continuous?\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & 
Not reliably, but after we have a better technology to manipulate the minimum units, we can predict the physical phenomenon and their computation model by a better discrete model. & 
Not reliably, because precision can never be high enough. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum models &  & Quantum Theories and Computing over Finite Fields & Quantum Interval-Valued Probability Measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison} with their
inspired quantum models. The physical quantity manipulated by the
quantum circuit model, the most widely used model of quantum computing,
is the probability amplitudes of states. \todo{Introduction Cleaned
up until here. }Because we are agnostic about whether probability
amplitude is quantized or can really be valued in the continuum of
complex numbers, we tried to develop two different types of models:
the quantum theories and computing over finite fields are discrete
models, and the quantum interval-valued probability measures incorporate
the idea of measurement precision into the quantum theory. To distinguish
the modified quantum theories from the usual ones, the usual ones
will be called the ``conventional quantum theory'' where necessary
and it will be reviewed in Chapter~\ref{chap:Conventional-Quantum-Theory}~\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}.

@@

\begin{tabularx}{\linewidth}{YYYYYYY}
\toprule 
\addlinespace
 & Conventional & Modal & Discrete (I) & Discrete (II) & QPMFF & QIVPM\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 & Chapter~\ref{chap:Conventional-Quantum-Theory} & Sec.~\ref{modalquantum} & Sec.~\ref{discretequantumtheoryI} to \ref{sec:Geometry-of-Entangled-States} & Sec.~\ref{sec:Discrete-Quantum-Theory-(II)} & Sec.~\ref{sec:Toward-IVPM} & Chapter~\ref{chap:QIVPM}\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
States space & $\mathbb{C}^{D}$ & $\Fq^{D}$ & $\Fpp^{D}$ & Local region in $\Fpp^{D}$ & $\Fpp^{D}$ & $\mathbb{C}^{D}$\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Likelihood of events is predicted by & Real-valued probability & Possible or impossible & Possible or impossible & Cardinal probability & Real-valued probability, but no sensible Born rule & Interval-valued probability\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Expectation value & Defined & Undefined & Undefined & Undefined &  & Defined\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 &  & Sec.~\ref{modalquantumcomputing} & Sec.~\ref{discretequantumcomputingI} & Sec.~\ref{discretequantumcomputingII} &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Hadamard transformation & Included & May not be included & Included & Included &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Solve $\usat$ efficiently & Unlikely & Yes & Partially & Unlikely &  & \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Grover search algorithm & Yes &  &  & Yes &  & \tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
\todo{Beginning of Merge Left}Since conventional probability amplitude
is in the field of complex numbers, and the field structure is essential
to compute the probability from some probability amplitudes, we in
Chapter~\ref{chap:DQT=000026DQC} tried to build discrete quantum
theories and computing by considering them over finite fields.
